{Gravitational wave sources in our Galactic backyard:}\\{Predictions for BHBH, BHNS and NSNS binaries detectable with LISA

### 2022-07-29 paper/tex/paper.tex
update floor citations
define referee command
clarify where stellar-mass sources come from
further updates after review

\documentclass[twocolumn, twocolappendix]{aastex63}
\received{\today}
\shorttitle{Detecting DCOs with LISA}
\graphicspath{{../figures/}}

@@

\newcommand{\citefloorp}{(\citealt{Broekgaarden+2021}, Broekgaarden et al.\ (in prep.))}

@@

\acknowledgments{We thank Katie Brevik, Will Farr, Rob Farmer, Valeriya Korol, Floris Kummer, Eva Laplace, Mike Lau, Tyson Littenberg, Ilya Mandel, Mathieu Renzo, Alberto Sesana, Katie Sharpe, Simon Stevenson, the CCA GW group and COMPAS collaboration for insightful discussions. TW also thanks Terence Lovatt for his advice and help on improving an earlier draft of this project. Portions of this study were inspired by the LISA Sprint at the Center for Computational Astrophysics of the Flatiron Institute, supported by the Simons Foundation. This project was funded in part by the National Science Foundation under Grant No. (NSF grant number 2009131), the European Union’s Horizon 2020 research and innovation program from the European Research Council (ERC, Grant agreement No. 715063), and by the Netherlands Organization for Scientific Research (NWO) as part of the Vidi research program BinWaves with project number 639.042.728. We further acknowledge the Black Hole Initiative funded by a generous contribution of the John Templeton Foundation and the Gordon and Betty Moore Foundation.}

\software{We used LEGWORK to evolve sources over time and calculate signal-to-noise ratios \citep{Wagg+2021}. It is freely available at \url{https://legwork.readthedocs.io/en/latest/}. Simulations in this paper made use of the COMPAS rapid binary population synthesis code \citep{COMPAS:2021methodsPaper}. The simulations performed in this work were simulated with a COMPAS version that predates the publicly available code. Our version is most similar to v02.13.01 of the publicly available COMPAS code. Requests for the original code can be made to Floor Broekgaarden. The authors used {\sc{STROOPWAFEL}} from \citep{Broekgaarden+2019}, publicly available at \url{https://github.com/FloorBroekgaarden/STROOPWAFEL}\footnote{For the latest pip installable version of STROOPWAFEL please contact Floor Broekgaarden.}.
The authors also made use of Python (v3.8), available at \url{http://www.python.org}. In addition the following Python packages were used: \texttt{matplotlib} \citep{Hunter+2007}, \texttt{NumPy} \citep{2020NumPy-Array}, \texttt{Astropy} \citep{AstropyCollaboration+2013,AstropyCollaboration+2018}, \texttt{Seaborn} \citep{Waskom+2021}, \texttt{SciPy} \citep{2020SciPy-NMeth}, \texttt{h5py} \citep{Collette+2021} and \texttt{Jupyter Lab} \citep{Kluyver+2016}. This research has made use of NASA’s Astrophysics Data System Bibliographic Services. We also made use of the computational facilities from the Harvard FAS Research Computing cluster.}

\bibliographystyle{aasjournal}
\bibliography{references}

>>>>

\documentclass[twocolumn, twocolappendix, linenumbers]{aastex63}
\received{\today}
\shorttitle{Detecting DCOs with LISA}
\graphicspath{{../figures/}}

@@

@@

\acknowledgments{We thank the anonymous referee for critically reviewing our initial version and providing many helpful comments. We thank Katie Brevik, Will Farr, Rob Farmer, Valeriya Korol, Floris Kummer, Eva Laplace, Mike Lau, Tyson Littenberg, Ilya Mandel, Mathieu Renzo, Alberto Sesana, Katie Sharpe, Simon Stevenson, the CCA GW group and COMPAS collaboration for insightful discussions. TW also thanks Terence Lovatt for his advice and help on improving an earlier draft of this project. Portions of this study were inspired by the LISA Sprint at the Center for Computational Astrophysics of the Flatiron Institute, supported by the Simons Foundation. This project was funded in part by the National Science Foundation under Grant No. (NSF grant number 2009131), the European Union’s Horizon 2020 research and innovation program from the European Research Council (ERC, Grant agreement No. 715063), and by the Netherlands Organization for Scientific Research (NWO) as part of the Vidi research program BinWaves with project number 639.042.728. We further acknowledge the Black Hole Initiative funded by a generous contribution of the John Templeton Foundation and the Gordon and Betty Moore Foundation.}

\software{We used LEGWORK to evolve sources over time and calculate signal-to-noise ratios \citep{LEGWORK_joss}. It is freely available at \url{https://legwork.readthedocs.io/en/latest/}. Simulations in this paper made use of the COMPAS rapid binary population synthesis code \citep{COMPAS:2021methodsPaper}. The simulations performed in this work were simulated with a COMPAS version that predates the publicly available code. Our version is most similar to v02.13.01 of the publicly available COMPAS code. Requests for the original code can be made to Floor Broekgaarden. The authors used {\sc{STROOPWAFEL}} from \citep{Broekgaarden+2019}, publicly available at \url{https://github.com/FloorBroekgaarden/STROOPWAFEL}\footnote{For the latest pip installable version of STROOPWAFEL please contact Floor Broekgaarden.}.
The authors also made use of Python (v3.8), available at \url{http://www.python.org}. In addition the following Python packages were used: \texttt{matplotlib} \citep{Hunter+2007}, \texttt{NumPy} \citep{2020NumPy-Array}, \texttt{Astropy} \citep{AstropyCollaboration+2013,AstropyCollaboration+2018}, \texttt{Seaborn} \citep{Waskom+2021}, \texttt{SciPy} \citep{2020SciPy-NMeth}, \texttt{h5py} \citep{Collette+2021} and \texttt{Jupyter Lab} \citep{Kluyver+2016}. This research has made use of NASA’s Astrophysics Data System Bibliographic Services. We also made use of the computational facilities from the Harvard FAS Research Computing cluster.}

\bibliographystyle{aasjournal}
\bibliography{paper}

### 2022-07-29 paper/tex/sections/conclusion.tex

We provide predictions for the detection rate and population properties of LISA detectable BHBH, BHNS and NSNS.
%
We use a novel empirically-informed analytical model for the metallicity dependent star formation history of the Milky Way, calibrated against the APOGEE stellar spectroscopic survey. We use this to model Monte-Carlo realisations of the present-day BHBH, BHNS and NSNS populations in our Milky Way. 
%
For the binary population, we use the results of a large grid of simulations performed with the rapid population synthesis code COMPAS\edit1{, which can reproduce the current constraints on the overall GW rates \citep{Broekgaarden+2021,Broekgaarden+2021b, Abbott+2021_GWTC3}}. Our suite of simulations consider 20 physics variations, which represent the most common uncertainties in binary physics.
%
We investigate the detectability of sources in the planned space-based GW detector LISA using the LEGWORK package \citep{Wagg+2021}, that we specifically developed for this purpose and is publicly available. 
%
We investigate the results expected for a 4- and 10-year LISA mission. Our main conclusions can be summarised as:
\begin{enumerate}
    \item \textbf{Total detections:} We predict \rangeFourYear{} detections in a 4-year LISA mission, across all our simulations for varying physics assumptions. This increases to about \rangeTenYear{} for a 10-year LISA mission. Although the number of detections per type can vary by about 2 orders of magnitude, we find that the total detection rate is fairly robust, among the variations we have considered (see Table~\ref{tab:detection_rates}).

>>>>

We provide predictions for the detection rate and population properties of LISA detectable BHBH, BHNS and NSNS.
%
We use a novel empirically-informed analytical model for the metallicity dependent star formation history of the Milky Way, calibrated against the APOGEE stellar spectroscopic survey. We use this to model Monte-Carlo realisations of the present-day BHBH, BHNS and NSNS populations in our Milky Way. 
%
For the binary population, we use the results of a large grid of simulations performed with the rapid population synthesis code COMPAS \citep{Broekgaarden+2021,Broekgaarden+2021b}. Our suite of simulations consider 20 physics variations, which represent the most common uncertainties in binary physics.
%
We investigate the detectability of sources in the planned space-based GW detector LISA using the LEGWORK package \citep{LEGWORK_apjs}, that we specifically developed for this purpose and is publicly available. 
%
We investigate the results expected for a 4- and 10-year LISA mission. Our main conclusions can be summarised as:
\begin{enumerate}
    \item \textbf{Total detections:} We predict \rangeFourYear{} detections in a 4-year LISA mission, across all our simulations for varying physics assumptions. This increases to about \rangeTenYear{} for a 10-year LISA mission. Although the number of detections per type can vary by about 2 orders of magnitude, we find that the total detection rate is fairly robust, among the variations we have considered (see Table~\ref{tab:detection_rates}).

### 2022-07-29 paper/tex/sections/discussion.tex

>>>>

\edit2{\paragraph{Correspondence with ground-based detections} The models used in this paper were shown to be consistent with the inferred merger rates from the GWTC-3 catalogue available at the time of submission (for example the BHBH rate was quoted as $16-130 \unit{Gpc^{-3}}{yr^{-1}}$ in section IV.A of version 1 and 2 on the arXiv in Nov 2021 \citealt{GWTC-3-previous}).
We note, however, that the upper limit on the rate has come down by a factor two since, after the completion and submission of our work. The inferred BHBH merger rate is now quoted as $16-61 \unit{Gpc^{-3}}{yr^{-1}}$ in the final version of the GWTC-3 paper that appeared in Feb 2022 \citep{Abbott+2021_GWTC3}. This means that some model variations may over predict the current rates.}

### 2022-07-29 paper/tex/sections/methods.tex

We use the grid of \nModels{} binary population synthesis simulations recently presented in \citet{Broekgaarden+2021,Broekgaarden+2021b}. This grid of simulations is synthesised using the rapid population synthesis code \href{https://compas.science}{COMPAS} \citep{Stevenson+2017,Vigna-Gomez+2018,Stevenson+2019,Broekgaarden+2019}. COMPAS follows the approach of the population synthesis code BSE \citep{Hurley+2000,Hurley+2002} and uses fitting formula and rapid algorithms to efficiently predict the final fate of binary systems. The code is open source and documented in the papers listed above, the online documentation\footnote{\url{https://compas.science}} and in the methods paper \citep{COMPAS:2021methodsPaper}. \edit1{The model we refer to as our fiducial model allows for BHs with masses in the heavily debated mass gap, but we also consider \nModels{} physics variations. \citet{Broekgaarden+2021b} showed that all of these physics variations are still consistent with current constraints on the overall inferred GW rates, at least when considering large variations of the assumed metallicity specific cosmic star formation rate \citep{Abbott+2021_GWTC3}}. We summarise the main assumptions and settings relevant for this work in Appendix~\ref{app:pop_synth}.

@@

\subsection{Gravitational wave detection}\label{sec:gw_detection}
We use the Python package \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} \citep{Wagg+2021} to evolve binaries and calculate their LISA detectability. For a full derivation of the equations given below see \citep[][Section~3]{Wagg+2021}, or the LEGWORK documentation \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{\faBook}.

@@

We use \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} \citep{Wagg+2021} to calculate the signal-to-noise ratio for each binary and the package ensures that enough harmonics are computed for each binary such that the error on the gravitational-wave luminosity remains below 1\%.

>>>>

We use the grid of \nModels{} binary population synthesis simulations recently presented in \citet{Broekgaarden+2021,Broekgaarden+2021b}. This grid of simulations is synthesised using the rapid population synthesis code \href{https://compas.science}{COMPAS} \citep{Stevenson+2017,Vigna-Gomez+2018,Stevenson+2019,Broekgaarden+2019}. COMPAS follows the approach of the population synthesis code BSE \citep{Hurley+2000,Hurley+2002} and uses fitting formula and rapid algorithms to efficiently predict the final fate of binary systems. The code is open source and documented in the papers listed above, the online documentation\footnote{\url{https://compas.science}} and in the methods paper \citep{COMPAS:2021methodsPaper}. \edit1{The model we refer to as our fiducial model allows for BHs with masses in the heavily debated mass gap, but we also consider \nModels{} physics variations. We summarise the main assumptions and settings relevant for this work in Appendix~\ref{app:pop_synth}. \edit2{We used these models since \citet{Broekgaarden+2021b} showed that each physics variation led to a rate prediction that was consistent with the constraints available from ground-based detections in GWTC-3 that were released at that time \citep{GWTC-3-previous}. We note that the upper limit on the rates has come down by a factor of two since the original appearance of this catalogue on the arXiv \citep{Abbott+2021_GWTC3}. We discuss this in further detail in Section~\ref{sec:caveats}}.}

@@

\subsection{Gravitational wave detection}\label{sec:gw_detection}
We use the Python package \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} \citep{LEGWORK_apjs} to evolve binaries and calculate their LISA detectability. For a full derivation of the equations given below see \citep[][Section~3]{LEGWORK_apjs}, or the LEGWORK documentation \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{\faBook}.

@@

We use \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} \citep{LEGWORK_apjs} to calculate the signal-to-noise ratio for each binary and the package ensures that enough harmonics are computed for each binary such that the error on the gravitational-wave luminosity remains below 1\%.
